GeckoGraph: A Visual Language for Polymorphic Types

### 2025-04-08 main.tex
keywords
changes from john

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

\end{keyword}

@@

\section{Introduction} \label{sec:intro}
In programming languages, a polymorphic type \cite{Cardelli1987-fp} can represent values of different types while providing a interface to describe common behaviors for those values. Polymorphic types are central to the succinctness of contemporary statically-typed functional languages, enabling a considerable degree of type-safe abstraction and hence reusability. Parametric polymorphism, a common technique to denote polymorphic type, is available in many programming languages, from functional languages such as Haskell and ML to imperative and multi-paradigm languages such as Rust\cite{Klabnik_undated-wx} and Go\cite{Griesemer_undated-ff}. Polymorphism allows programs to be written in a way that is more generic and adaptable to different data types, enabling greater flexibility and code reuse. Polymorphic types are ideal for modeling abstractions, such as properties of mathematical objects and laws that hold on these objects.

@@

While type polymorphism is one of the oldest topics in programming language theory \cite{Cardelli1987-fp}, little research focuses on the \textit{usability} of polymorphic types. It is often argued that the expressiveness and power of type systems often come at the cost of usability~\cite{Hage2020-hg}. We aim to investigate the challenges of using polymorphic types and explore how to improve their usability with visualization and modern HCI techniques. In this paper, we propose GeckoGraph, a graphical notation for types. GeckoGraph aims to complement traditional text-based type notation and make reading, understanding, and comparing types easier. GeckoGraph is prototyped and verified iteratively, leading to a design with visual clarity applicable to many programming contexts. Our study evaluating 714 participants' ability to solve a series of programming challenges using GeckoGraph versus text-based type annotation is, to our knowledge, the largest controlled user study of a functional programming tool or concept. The study results show that GeckoGraph helps improve programmers' ability to succeed in resolving the type challenges, especially for novice programmers.

@@

GeckoGraph is a visual notation for type annotations in statically typed programming languages. It is intended to work tangibly with text-based annotations, but uses colors, shapes, and symbols to make structures of types easy to identify at a glance. In this section, we describe the design of GeckoGraph and highlight some unique benefits of programming with GeckoGraph.

@@

\paragraph{\textbf{(D3) Optimized for comparing and searching}} \label{goal3} GeckoGraph should aim to make the task of comparing two types easy, especially to make subtle differences in text-based notation harder to miss. This also includes the task of choosing an ideal function from a list of potential functions. For example, programmers search for a desired function from a documentation site with only partial knowledge of its type (e.g., the arity of a function, containing a known type, or type classes requirements).

@@

Curried functions (e.g., \texttt{ a -> b -> c}) display as two cells of functions merged together \includegraphics[height=1em]{figures/Curry.png}. It displays as if the second function covers the right half of the first function, indicating that the second function is the return type of the first. Regular high-order functions (e.g., \texttt{(a -> b) -> c} ) follow the rules of functions and nested data types \includegraphics[height=1.2em]{figures/HOF.png}. The placement of function indicators aims to make it easy to find desired functions in the documentation site based on function arity and high-order functions (design goal \ref{goal3}). It is easy to tell high-order functions from the vertical position of its function indicator. Similarly, it is easy to count the arity of a function by counting the number of horizontally connected function indicators (Fig. \ref{fig:indicator}).

\begin{figure}[]
  \includegraphics[width=\linewidth]{figures/Indicator}
  \caption{
        \label{fig:indicator}
        An example of visual hirachy of the function indicator. The function indicator can be used to easily identify the arity of a function type by counting the connecting function indicators. For higher-order functions where functions are arguments or return values of other functions, it is very easy to see the ``level" of functions. 
  }
\end{figure}

\paragraph{Type Classes} 
Type classes are an intrinsic part of Haskell \cite{Hudak2007-kn}, and many other functional languages. In GeckoGraph, the type classes (e.g., \texttt{ (A a, B a) => a}) are indicated in the extended area below one or more GeckoGraph cells \includegraphics[height=1.2em]{figures/TypeClass.png}. Each type class required on a type variable displays as a square indicator aligned on the right of the extended area. In GeckoGraph, a type-class gliph is embeded in every type variable that requires this  class. This means when displaying the type \texttt{(==) :: Eq a => 
a -> a -> Bool} in GeckoGraph, the constraint \texttt{Eq} appears in twice on both occurrences of the type variable \texttt{a}.

The GeckoGraph type class's design promotes the type class placements rather than the type class names. Programmers can easily see where and how many type classes are required, but they may need an extra step (global legends of the color mapping or pop-up window) to identify the name of the type class. We believe that this is a well justified trade-off. The strength of this appraoch lies in situations when type class context is complex. When reading a type \texttt{(A a, A c, B a, B b, C b) => a -> b -> c}, programmers may need to switch back and forth to remember which type classes are needed on which variable. GeckoGraph uses approximity to help minimize the effort to associate each type variable with all its type classes (design goal \ref{goal2}).

\subsection{Strengths of GeckoGraph}\label{sec:benefits}

\paragraph{Generalization patterns in type classes}
A common source of confusion among novice Haskell programmers is distinguishing between functions that operate on generalized types and those that work with concrete types \cite{Tirronen2015-nr}. For instance, the type constraints {\tt Integral a => a} and {\tt Integer}, {\tt Floating a => a} and {\tt Float}, as well as {\tt Foldable} instances versus {\tt []} can be particularly challenging. Many Haskell learners may find the differences between concrete types and their generic counterparts irrelevant to their programming tasks, thereby complicating their quest for solutions. In the example shown in Fig. \ref{fig:example-foldable}, a programmer mistakenly supplies a \texttt{Char} instead of a \texttt{String} literal. The compiler responds with an error message that compares the provided \texttt{Char} type to a confusingly named type \texttt{t0 a0}. In this case, the information has been generalized too much and the programmer can hardly find their bearing to the original problem. This level of generalization can undermine a programmer's confidence in understanding the language and navigating type errors. In GeckoGraph, both list types and types with a Foldable instance share a similar structure. Therefore, the type \texttt{t0 a0} in the error message matches the shape of \texttt{[a]}. Bring programmer closer to the real issue, trying to match a complex type with a concrete simple type. This feature lowers the entry barrier for languages like Haskell, where even basic arithmetic functions {\tt +} and {\tt -} are defined on generic types. This approach lowers the barrier to introduce foundational language concepts without sacrificing generality.

\paragraph{Consistent color scheme}\label{par:color-scheme}
A common task in programming is to scan for a desired function from a sea of potentially useful functions, such as library documentation. During scanning, programmers often have partial knowledge of the desired function, e.g., the arity, one of the argument types, or the type class it must fulfill.  A typical example is conversion: using a known \texttt{String} type to produce a desired \texttt{Data.Text}  type. Another example is the `lookup' function: using a known \texttt{Data.Map a b} to produce a desired \texttt{b} type. GeckoGraph supports this task by using consistent colors for the same type identifier. Programmers can rely on the color grouping to scan for the desired type in their project or in third-party library documentation.

\paragraph{Advanced Type Feature Visualization}
The design of GeckoGraph enables the visualization of many advanced type-level features. \textbf{Kind visualization}: Kind system \cite{Jones1995-lm} is used in Haskell to indicate whether a type is type constructor (e.g. [], Maybe) or a type constant (e.g. Bool, Char), because of this, kind is often described as the type of types. Goanna infer the kind of type variables,  consistently display this information in GeckoGraph. For example, in figure \ref{fig:advanced} (A), the variable \texttt{a} is infered to have the kind \texttt{* -> *} because of its use on the right-hand side. GeckoGraph respects this kind information and displays the same kind level on the first occurance of \texttt{a}, but over an empty structure, indicated using a dotted outline. \textbf{Qualified constraints}: GeckoGraph's type class notation naturally extends to support qualified constraints. In the type \texttt{forall b. A (a b) => a b}, GeckoGraph shows the scope type class requirement on \texttt{a b} (Fig. \ref{fig:advanced} B).
\textbf{Multiple Parameter Type Class}:  GeckoGraph supports multiple parameter type classes by using multiple shapes with the same color hue to indicate the different parameters of the same type class. For example, for the type \texttt{A a b => a b},  GeckoGraph shows that the variables a and b both need an A class, but they are the different parameters of A (Fig. \ref{fig:advanced} C).

\paragraph{Precise Interactivity}
Modern programming environments often allow programmers to mouse over part of the source code to query detailed information, such as definition, references or documentation. However, with text-based source code, it is often hard to distinguish whether programmers want the most specific fragment under the cursor or larger blocks. Because of its graphical layout, GeckoGraph allows programmers to precisely select which part of a type signature they intend to query, that is, in Fig. \ref{fig:advanced} (D) when the user mouses over the type class box (orange square) under the second occurrence of \texttt{a} the type class it represents is revealed in detail, a sepecific query that is hard to hard to execute with point and click over text based type annotation.

@@

\paragraph{Language Agnostic}
GeckoGraph can be implemented in any language that exercises static typing. We identified applications of this feature in programming and education. In programming projects often use more than one programming languages, situations include projects using foreign function interfaces or multiple languages for client-side and server-side programming. In these cases, GeckoGraph provides a shared notation to describe the functionality and features of systems, dispide their text-based notations are different. In teaching and learning programming languages, GeckoGraph removes the nomenclature difference in different programming languages. For example, when describing the product type in algebraic data types, different language communities use various names: record, named tuple, struct, etc. GechoGraph helps to unify the same programming concept and ignore the minute nomenclature differences .

\subsection{Previous iterations of GeckoGraph}
GeckoGraph was designed through many different iterations. Many research methods were used to verify ideas, including prototyping, cognitive walk-throughs, and formative studies.  We list some notable design elements and their major feedback.

\paragraph{\textbf{Encoding the direction of function types}}
This can be found in the design v1 (Fig. \ref{fig:previous}). This design shares many similarities with a prior project \cite{Jung2000-oc} in type visualization. Although many students and experts agree on this, its use of space on both width and height scales is proportional to the types in question, causing too much inconvenience. In addition, this design does not produce a canonical form for a curried function. For example, for composition (.), it is unclear whether it takes two functions as input and returns a binary function or two functions and a single structure as input and returns a different structure.

\paragraph{Encoding the depth and size} Grayscales, gradience, size, and simulated 3-dimensional elevation are promising visual representations for depicting numeric dimensions such as the depth and width of the parse tree. However, gradient and elevation were dismissed because of their requirements for a more demanding rendering process, making GeckoGraph harder to implement in more restricted user interfaces, such as the command line. In addition, all of these visual dimensions reduce readability when scaled down to a very small size.

\paragraph{Encoding symbolic names} In some variations, we tested using different geometric shapes to indicate the symbolic names of type variables and concrete types. We decided against using icons due to the limited number of different shapes until they were indistinguishable. The color provides more encoding spaces, and the letters provide familiarity with the original type annotations. This was shown to help reduce friction in the adoption of GeckoGraph.

@@

At each level of the game, the programmer must select the right set of functions, and compose them correctly to achieve the target result. Not all provided functions are useful. In particular, for this level, participants must discover that only \texttt{f4} and \texttt{f2} are necessary to produce the desired results. One implementation that satisfies the target type is \texttt{zeroTohero z = f2 (f4  z)}.

@@

The experiment uses a between-subject design. Participants are assigned to one of two groups. The tasks asiigned to the two groups are the same order. Participants receive both treatments (with and without GeckoGraph) during their runs but in different order.   Group one is assisted by GeckoGraph on even levels and text-only type annotation on odd levels. Group two participants are the same but with the order flipped. The number of participants in the two groups is counterbalanced.

@@

\subsection{Task Design} \label{subsection:task}
During the study, participants receive 10 tasks, represented by the 10 levels of the game {\it zeroToHero}. The tasks start off easy but gradually increase in difficulty.  In each task, a target type of the function \texttt{zeroToHero} is shown to the participants. Participants are provided with a list of available functions to implement the target function, and to ensure it assume the target type. This is to simulate the tasks of selecting useful functions from a library. Participants are not allowed to use any other functions or variables outside the provided functions; even the Haskell prelude is not available. This ensures that everyone has the same knowledge and minimizes the effect of familiarity.

@@

The 10 levels are designed to gradually increase difficulty. From the results of the experiment, most of the tasks align with this trend. However, three tasks stand out in Fig. \ref{fig:level-time}.  Level 7 (mean = 334 seconds) is the hardest task in the game in terms of time, followed by level 8 (mean = 228 seconds) and level 5 (mean = 224 seconds). To complete an average level, the beginner group uses an average of 100 seconds, the familiar group uses 90 seconds, the knowledgeable group uses 80 seconds, and the expert group uses 70 seconds. This roughly aligns with self-reported expertise. We show that the task time on each level follows normal distributions using a Shapiro-Wilk test \cite{Shaphiro1965-dx} (p-value  $ \leq 1.018 \times 10^-16$, for an alpha value of 0.05, p less than 0.05 is considered normal distribution).

Level 5, 7 and 8 are the only three levels that include functions from standard Haskell library, baring the \texttt{(.)} and \texttt{(\$)} provided for convenience. Level 3 requires programmers to use the \texttt{fst} and \texttt{snd} functions to extract value from a tuple. Level 7 requires programmers to use the \texttt{(<*>)} function of the \texttt{Applicative} class, while level 8 the \texttt{fmap} function of the \texttt{Functor} class. The authors speculate that the more experienced participants are much more familiar with these functions, hence the strong contrast on these three levels.

@@

\subsection{Success rate}
We saw that, overall, GeckoGraph provides a higher success rate (96.88\%) than text-based type notation (94.62\%). This trend can be seen in every experienced group: beginner group (95. 12\% vs. 92. 68\%), familiar group (97. 39\% vs. 93. 34\%), knowledgeable group (96. 82\% vs. 96. 06\%) and expert group (98. 2\% vs. 96. 40\%). We saw the significance decrease as the user's experience increased. When performing a proportion test on each group, we see that the effect is most significant with the beginner group and reject the null hypothesis (z score = 2.0228, p-value = 0.0431), followed by the familiar group (z score = 1.7495, p-value = 0.0802). The knowledgeable group (z score = 1.0295, p-value = 0.3032) and the expert group (z score = 0.8660, p-value = 0.3756) show less significant differences between treatments.

@@

Level 10 is the most cognitively demanding: the provided functions are the lengthiest, and the solution is challenging and deceptive. Programmers must use the provided rotate function {\it f2} to adjust the positions of the four components of the {\it Zero} value four time to get to the correct position. Completing the level requires applying the functions {\it f1}, {\it f2}, and {\it f3} in a nested sequence that is 8 layers deep. Consequently, programmers need to frequently switch contexts to recall the purpose of each function. We speculate that this level demands the mental flexibility typical of a Haskell expert to quickly internalize these functions. GeckoGraph helps less experienced programmers to achieve the same level of comprehension fluency by adding structual clarity.

@@

\subsection{Qualitative Feedback}
In their responses to the post-study survey, most programmers believe that the design of GeckoGraph is intuitive and that its appearance in the interface does not cause distraction.
For the question ``Do you find the GeckoGraph distracting", most of the participants rated a negative score, with an average of 2.88 (Fig. \ref{fig:qualitative} left). For the question ``How intuitive do you find the GeckoGraph?", we saw a reverse correlation of experience (Fig. \ref{fig:qualitative} Middle): experts find the GeckoGraph most intuitive (5.07), followed by the knowledgeable group (4.87), and the familiar group (4.80). The beginner group found it to be the least intuitive but still rated a positive score of (4.71). 
When answering the question ``How helpful do you find GeckoGraph in finding the solution during the game?", the answer is more divided into different experience groups (Fig. \ref{fig:qualitative} right). It is slightly positive for beginners (4.25) and slightly negative for the other groups, the familiar group (3.86) and the knowledgeable group (3.32). The expert group finds that GeckoGraph is relatively unhelpful, with an average score of 2.95.

Two interesting trends emerge. First, the more experienced the users are, the more likely they find GeckoGraph intuitive and easy to use. We believe this aligns with our design decision to make GeckoGraph resemble text-based type annotation. Those familiar with text-based type annotation find GeckoGraph second nature to read. Second, the correlation reverses when it comes to perceived helpfulness. The more experienced the users are, the less likely they are to find GeckoGraph helpful. We speculate this pattern occurs because users who are experts in text-based type annotations naturally prefer reading text-based notation and tend to ignore GeckoGraph, diminishing its usefulness. Conversely, novice users are more likely to be drawn to GeckoGraph and find its visualization offers more clarity and useful insights.

@@

\paragraph{Task design}
In our human study, most of the provided functions are very abstract. These functions are created by the authors solely for the gamified study. They are designed to be different from well-known Haskell functions to minimize the familiarity variable. They are also designed with an interest of being puzzling and fun.  These functions may not be the most representative of real-world Haskell programming.

@@

\subsubsection{Identify the Most Important Features}
One trend that we saw from the qualitative feedback is that programmers find GeckoGraph helpful for finding patterns and important features of the types. Programmers are very positive about GeckoGraph's ability to reveal the most helpful features of a type in distinctive visual elements such as color, length, and height.

\textbf{The colors of GeckoGraph} help programmers to see the permutation of type variables in the input and output of a function. A recent review \cite{Zeng2023-jz} of 59 graphical perception articles showed that combining solid color hue in a filled shape provides stronger visual perception for nominal data such as type identifiers. One example of GeckoGraph's effective use of color is the ``rotation" function in the user study (Fig. \ref{fig:rotate}). With text-based type notation, programmers often rely on mnemonic devices such as alphabetic ordering or naming conventions. For example, the rotation function \texttt{f2 :: Zero a b c d -> Zero b c d a} in the game is less recognizable if changed to \texttt{f2 :: Zero e v m h -> Zero v m h e}. To quote a participant, ``GeckoGraph is quite intuitive to see the permutations of the arguments. Also, to see how to produce and consume arguments."

\textbf{The horizontal axis of GeckoGraph} often becomes intuitive when identifying differences in function arities. For example, in Fig. \ref{fig:add3}, the programmer intended to implement a function that sums 3 integers. In the implementation, the programmer missed a \texttt{(+)} function at the end; the resulting function type is largely different in length. It is also clear that the function needs to apply to one more binary function to satisfy the length requirement.

\begin{figure}[]
  \includegraphics[width=0.6\linewidth]{figures/Add3}
  \caption{\label{fig:add3} An implementation of function \texttt{add3} but the author missed an (+) from the correct implementation (.) ((+) .) (+). GeckGraph highlights the difference in arity, and reveals that a binary function is needed on the right-hand side for the arity to match. }
\end{figure}

>>>>

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)
Polymorphic Types \sep Functional Programming \sep Visual Languages \sep Visualization
\end{keyword}

@@

\section{Introduction} \label{sec:intro}
In programming languages, a polymorphic type \cite{Cardelli1987-fp} can represent values of different types while providing an interface to describe common behaviors for those values. Polymorphic types are central to the succinctness of contemporary statically-typed languages, enabling a considerable degree of type-safe abstraction and reusability. Parametric types are available in many programming languages, from functional languages such as Haskell and ML to imperative and multi-paradigm languages such as Rust\cite{Klabnik_undated-wx} and Go\cite{Griesemer_undated-ff}. Polymorphism allows programs to be written in a way that is more generic and adaptable to different data types, enabling greater flexibility and code reuse. Polymorphic types are ideal for modeling abstractions, such as properties of mathematical objects and laws that hold on these objects.

@@

Polymorphic type is one of the oldest topics in programming language theory \cite{Cardelli1987-fp}. However, little research focuses on the \textit{usability} of polymorphic types. It is often argued that the expressiveness and power of type systems often come at the cost of usability~\cite{Hage2020-hg}. We aim to investigate the challenges of using polymorphic types and explore how to improve their usability with visualization and modern HCI techniques. In this paper, we propose GeckoGraph, a graphical notation for types. GeckoGraph aims to complement traditional text-based type notation and make reading, understanding, and comparing types easier. GeckoGraph is prototyped and verified iteratively, leading to a design with visual clarity applicable to many programming contexts. Our study evaluating 714 participants' ability to solve a series of programming challenges using GeckoGraph versus text-based type annotation is, to our knowledge, the largest controlled user study of a functional programming tool or concept. The study results show that GeckoGraph helps improve programmers' ability to succeed in resolving the programming challenges, especially for novice programmers.

@@

GeckoGraph is a visual notation for type annotations in statically typed programming languages. It is intended to work tangibly with text-based annotations but uses colors, shapes, and symbols to make structures of types easy to identify at a glance. In this section, we describe the design of GeckoGraph and highlight some unique benefits of programming with GeckoGraph.

@@

\paragraph{\textbf{(D3) Optimized for comparing and searching}} \label{goal3} GeckoGraph aims to make the task of comparing two types easy, especially to make subtle differences in text-based notation harder to miss. This also includes the task of choosing an ideal function from a list of potential functions. For example, programmers search for a desired function from a documentation site with only partial knowledge of its type (e.g., the arity of a function, containing a known type, or type class requirements).

@@

Curried functions (e.g., \texttt{ a -> b -> c}) are displayed as two cells of functions merged together \includegraphics[height=1em]{figures/Curry.png}. It appears as if the second function covers the right half of the first function, indicating that the second function is the return type of the first. Regular high-order functions (e.g., \texttt{(a -> b) -> c} ) follow the rules of functions and nested data types \includegraphics[height=1.2em]{figures/HOF.png}. The placement of function indicators aims to make it easy to find desired functions in the documentation site based on function arity and high-order functions (design goal \ref{goal3}). It is easy to tell high-order functions from the vertical position of its function indicator. Similarly, it is easy to count the arity of a function by counting the number of horizontally connected function indicators (Fig. \ref{fig:indicator}).

\begin{figure}[]
  \includegraphics[width=\linewidth]{figures/Indicator}
  \caption{
        \label{fig:indicator}
        An example of visual hierarchy of the function indicator. The function indicator can be used to easily identify the arity of a function type by counting the connecting function indicators. For higher-order functions where functions are arguments or return values of other functions, it is very easy to see the ``level" of functions. 
  }
\end{figure}

\paragraph{Type Classes} 
Type classes are an intrinsic part of Haskell \cite{Hudak2007-kn}, and many other functional languages. In GeckoGraph, the type classes (e.g., \texttt{ (A a, B a) => a}) are indicated in the extended area below one or more GeckoGraph cells \includegraphics[height=1.2em]{figures/TypeClass.png}. Each type class required on a type variable displays as a square indicator aligned on the right of the extended area. In GeckoGraph, a type-class gliph is embedded in every type variable that requires this class. This means when displaying the type \texttt{(==) :: Eq a => 
a -> a -> Bool} in GeckoGraph, the constraint \texttt{Eq} appears in twice on both occurrences of the type variable \texttt{a}.

The GeckoGraph type class's design promotes the type class placements rather than the type class names. Programmers can easily see where and how many type classes are required, but they may need an extra step (global legends of the color mapping or pop-up window) to identify the name of the type class. We believe that this is a well-justified trade-off. The strength of this approach lies in situations when type class context is complex. When reading a type \texttt{(A a, A c, B a, B b, C b) => a -> b -> c}, programmers may need to switch back and forth to remember which type classes are needed on which variable. GeckoGraph uses approximity to help minimize the effort to associate each type variable with all its type classes (design goal \ref{goal2}).

\subsection{Key Design Criteria of GeckoGraph}\label{sec:benefits}

GeckoGraph was designed through many different iterations. Many research methods were used to verify ideas, including prototyping, cognitive walk-throughs, and formative studies. We discuss some notable visual dimensions in GeckoGraph and their justifications.

\paragraph{Visual Dimensions of GeckoGraph}
\textbf{The colors of GeckoGraph} help programmers to see the permutation of type variables in the input and output of a function. A recent review \cite{Zeng2023-jz} of 59 graphical perception articles showed that combining solid color hue in a filled shape provides stronger visual perception for nominal data such as type identifiers. One example of GeckoGraph's use of color is the ``rotation" function  (Fig. \ref{fig:rotate}) in the user study (Section \ref{sec:evaluation}). With text-based type notation, programmers often rely on mnemonic devices such as alphabetic ordering or naming conventions. For example, the rotation function \texttt{f2 :: Zero a b c d -> Zero b c d a} in the game is less recognizable if changed to \texttt{f2 :: Zero e v m h -> Zero v m h e}.

\textbf{The horizontal axis of GeckoGraph} often are helpful in identifying differences in function arities. For example, in Fig. \ref{fig:add3}, the programmer intended to implement a function that sums 3 integers. In the implementation, the programmer missed a \texttt{(+)} function at the end; the resulting function type is largely different in length. It is also clear that the function needs to apply to one more binary function to satisfy the length requirement.

\begin{figure}[]
  \includegraphics[width=0.6\linewidth]{figures/Add3}
  \caption{\label{fig:add3} An implementation of function \texttt{add3} but the author missed an (+) from the correct implementation (.) ((+) .) (+). GeckGraph highlights the difference in arity and reveals that a binary function is needed on the right-hand side for the arity to match. }
\end{figure}

\textbf{The vertical axis of GeckoGraph} often helps identify the number of levels in a nested complex type. This can often be very useful when inspecting mismatching type errors where data types are nested in different levels. Common examples include when programmers forget to apply the value to ``return" in a monadic block or to use \texttt{liftIO} to cast an \texttt{IO} effect. In Fig. \ref{fig:maybe}, the uses of \texttt{return} are excessive. It can be easily identified by examining the difference in the vertical layers of the two types. In text-based type notation, this is distinguished by different pairs of parentheses. However, parenthesis is an overloaded syntax in type notation. In Haskell, parentheses are used to enclose tuples \texttt{(a, b)}, specify the fixity \texttt{ (a -> b) -> c}, or have no effect \texttt{a -> (b -> c)}.

\paragraph{Support Advanced Type Features}
The design of GeckoGraph enables the visualization of many advanced type-level features. \textbf{Kind visualization}: Kind system \cite{Jones1995-lm} is used in Haskell to indicate whether a type is type constructor (e.g. [], Maybe) or a type constant (e.g. Bool, Char), because of this, kind is often described as the type of types. GeckoGraph infers the {\it kind} of type variables and consistently displays this information in GeckoGraph. For example, in figure \ref{fig:advanced} (A), the variable \texttt{a} is inferred to have the kind \texttt{* -> *} because of its use on the right-hand side. GeckoGraph respects this {\it kind} information and displays the same kind level on the first occurrence of \texttt{a}, but over an empty structure, indicated using a dotted outline. \textbf{Qualified constraints}: GeckoGraph's type class notation naturally extends to support qualified constraints. In the type \texttt{forall b. A (a b) => a b}, GeckoGraph shows the scope type class requirement on \texttt{a b} (Fig. \ref{fig:advanced} B).
\textbf{Multiple Parameter Type Class}:  GeckoGraph supports multiple parameter type classes by using multiple shapes with the same color hue to indicate the different parameters of the same type class. For example, for the type \texttt{A a b => a b},  GeckoGraph shows that the variables a and b both need an A class, but they are the different parameters of A (Fig. \ref{fig:advanced} C).

\paragraph{Provide Precise Interactivity}
Modern programming environments often allow programmers to mouse over part of the source code to query detailed information, such as definition, references, or documentation. However, with text-based source code, it is often hard to distinguish whether programmers want the most specific fragment under the cursor or larger blocks. Because of its graphical layout, GeckoGraph allows programmers to precisely select which part of a type signature they intend to query, that is, in Fig. \ref{fig:advanced} (D) when the user mouses over the type class box (orange square) under the second occurrence of \texttt{a} the type class it represents is revealed in detail, a sepecific query that is hard to hard to execute with point and click over text based type annotation.

@@

@@

At each level of the game, the programmer must select the right set of functions and compose them correctly to achieve the target result. Not all provided functions are useful. In particular, for this level, participants must discover that only \texttt{f4} and \texttt{f2} are necessary to produce the desired results. One implementation that satisfies the target type is \texttt{zeroTohero z = f2 (f4  z)}.

@@

The experiment uses a between-subject design. Participants are assigned to one of two groups. The tasks assigned to the two groups are in the same order. Participants receive both treatments (with and without GeckoGraph) during their runs but in a different order.   Group one is assisted by GeckoGraph on even levels and text-only type annotation on odd levels. Group two participants are the same but with the order flipped. The number of participants in the two groups is counterbalanced.

@@

\subsection{Task Design} \label{subsection:task}
During the study, participants receive 10 tasks, represented by the 10 levels of the game {\it zeroToHero}. The tasks start off easy but gradually increase in difficulty.  In each task, a target type of the function \texttt{zeroToHero} is shown to the participants. Participants are provided with a list of available functions to implement the target function and to ensure it assumes the target type. This is to simulate the tasks of selecting useful functions from a library. Participants are not allowed to use any other functions or variables outside the provided functions; even the Haskell prelude is not available. This ensures that everyone has the same knowledge and minimizes the effect of familiarity.

@@

The 10 levels are designed to increase difficulty gradually. From the results of the experiment, most of the tasks align with this trend. However, three tasks stand out in Fig. \ref{fig:level-time}.  Level 7 (mean = 334 seconds) is the hardest task in the game in terms of time, followed by level 8 (mean = 228 seconds) and level 5 (mean = 224 seconds). To complete an average level, the beginner group uses an average of 100 seconds, the familiar group uses 90 seconds, the knowledgeable group uses 80 seconds, and the expert group uses 70 seconds. This roughly aligns with self-reported expertise. We show that the task time on each level follows normal distributions using a Shapiro-Wilk test \cite{Shaphiro1965-dx} (p-value  $ \leq 1.018 \times 10^-16$, for an alpha value of 0.05, p less than 0.05 is considered normal distribution).

Levels 5, 7, and 8 are the only three levels that include functions from standard Haskell library, baring the \texttt{(.)} and \texttt{(\$)} provided for convenience. Level 3 requires programmers to use the \texttt{fst} and \texttt{snd} functions to extract value from a tuple. Level 7 requires programmers to use the \texttt{(<*>)} function of the \texttt{Applicative} class, while level 8 requires the \texttt{fmap} function of the \texttt{Functor} class. The authors speculate that the more experienced participants are much more familiar with these functions, hence the strong contrast on these three levels.

@@

\subsection{Success rate}
We saw that, overall, GeckoGraph provides a higher success rate (96.88\%) than text-based type notation (94.62\%). This trend can be seen in every experienced group: beginner group (95. 12\% vs. 92. 68\%), familiar group (97. 39\% vs. 93. 34\%), knowledgeable group (96. 82\% vs. 96. 06\%), and expert group (98. 2\% vs. 96. 40\%). We saw the significance decrease as the user's experience increased. When performing a proportion test on each group, we see that the effect is most significant with the beginner group and reject the null hypothesis (z score = 2.0228, p-value = 0.0431), followed by the familiar group (z score = 1.7495, p-value = 0.0802). The knowledgeable group (z score = 1.0295, p-value = 0.3032) and the expert group (z score = 0.8660, p-value = 0.3756) show less significant differences between treatments.

@@

Level 10 is the most cognitively demanding: the provided functions are the lengthiest, and the solution is challenging and deceptive. Programmers must use the provided rotate function {\it f2} to adjust the positions of the four components of the {\it Zero} value four times to get to the correct position. Completing the level requires applying the functions {\it f1}, {\it f2}, and {\it f3} in a nested sequence that is 8 layers deep. Consequently, programmers need to frequently switch contexts to recall the purpose of each function. We speculate that this level demands the mental flexibility typical of a Haskell expert to internalize these functions quickly. GeckoGraph helps less experienced programmers to achieve the same level of comprehension fluency by adding structural clarity.

@@

\subsection{Qualitative Feedback}
In their responses to the post-study survey, most programmers believe that the design of GeckoGraph is intuitive and that its appearance in the interface does not cause distraction.
For the question ``Do you find the GeckoGraph distracting", most of the participants rated a negative score, with an average of 2.88 (Fig. \ref{fig:qualitative} left). For the question ``How intuitive do you find the GeckoGraph?", we saw a reverse correlation of experience (Fig. \ref{fig:qualitative} Middle): experts find the GeckoGraph most intuitive (5.07), followed by the knowledgeable group (4.87), and the familiar group (4.80). The beginner group found it to be the least intuitive but still rated a positive score of 4.71. 
When answering the question ``How helpful do you find GeckoGraph in finding the solution during the game?", the answer is more divided into different experience groups (Fig. \ref{fig:qualitative} right). It is slightly positive for beginners (4.25) and slightly negative for the other groups, the familiar group (3.86) and the knowledgeable group (3.32). The expert group finds that GeckoGraph is relatively unhelpful, with an average score of 2.95.

Two interesting trends emerge. First, the more experienced the users are, the more likely they are to find GeckoGraph intuitive and easy to use. This aligns with our design decision to make GeckoGraph resemble text-based type annotation. Those familiar with text-based type annotation find GeckoGraph second nature to read. Second, the correlation reverses when it comes to perceived helpfulness. The more experienced the users are, the less likely they are to find GeckoGraph helpful. We speculate this pattern occurs because users who are experts in text-based type annotations naturally prefer reading text-based notation and tend to ignore GeckoGraph, diminishing its usefulness. Conversely, novice users are more likely to be drawn to GeckoGraph and find that its visualization offers more clarity and useful insights.

% To quote some feedback from participants: ``Types are much easier to understand by the GeckoGraph than by trying to parse all parentheses and understand the types from the signature. " ``It makes it easier to see at a glance when your output type is correct or what the difference between the current type and the target is."

% To quote a participant, ``GeckoGraph is quite intuitive to see the permutations of the arguments. Also, to see how to produce and consume arguments."

@@

\paragraph{Task design}
In our human study, most of the provided functions are very abstract. These functions are created by the authors solely for the gamified study. They are designed to be different from well-known Haskell functions to minimize the familiarity variable. They are also designed with an interest in being challenging and fun. These functions may not be the most representative of real-world Haskell programming.

@@

\subsubsection{Identify the Most Important Features}
One trend we saw from the qualitative feedback is that programmers find GeckoGraph helpful for finding patterns and important features of the types. Programmers are very positive about GeckoGraph's ability to reveal the most helpful features of a type in distinctive visual elements such as color, length, and height.
To quote some feedback from participants: ``{\it Types are much easier to understand by the GeckoGraph than by trying to parse all parentheses and understand the types from the signature.} " ``{\it It makes it easier to see at a glance when your output type is correct or what the difference between the current type and the target is.}"

\subsubsection{Generalize patterns in type classes}
A common source of confusion among novice Haskell programmers is distinguishing between functions that operate on generalized types and those that work with concrete types \cite{Tirronen2015-nr}. For instance, the type constraints {\tt Integral a => a} and {\tt Integer}, {\tt Floating a => a} and {\tt Float}, as well as {\tt Foldable} instances versus {\tt []} can be particularly challenging. Many Haskell learners may find the differences between concrete types and their generic counterparts irrelevant to their programming tasks, thereby complicating their quest for solutions. In the example shown in Fig. \ref{fig:example-foldable}, a programmer mistakenly supplies a \texttt{Char} instead of a \texttt{String} literal. The compiler responds with an error message that compares the provided \texttt{Char} type to a confusingly named type \texttt{t0 a0}. In this case, the information has been generalized too much and the programmer can hardly find their bearing to the original problem. This level of generalization can undermine a programmer's confidence in understanding the language and navigating type errors. In GeckoGraph, both list types and types with a Foldable instance share a similar structure. Therefore, the type \texttt{t0 a0} in the error message matches the shape of \texttt{[a]}. Bring the programmer closer to the real issue, trying to match a complex type with a concrete simple type. This feature lowers the entry barrier for languages like Haskell, where even basic arithmetic functions {\tt +} and {\tt -} are defined on generic types. This approach lowers the barrier to introducing foundational language concepts without sacrificing generality.

\subsubsection{Consistent color scheme}\label{par:color-scheme}
A common task in programming is to scan for a desired function from a sea of potentially useful functions, such as library documentation. During scanning, programmers often have partial knowledge of the desired function, e.g., the arity, one of the argument types, or the type class it must fulfill.  A typical example is conversion: using a known \texttt{String} type to produce a desired \texttt{Data.Text}  type. Another example is the `lookup' function: using a known \texttt{Data.Map a b} to produce a desired \texttt{b} type. GeckoGraph supports this task by using consistent colors for the same type identifier. Programmers can rely on the color grouping to scan for the desired type in their project or in third-party library documentation.
